A Formal Treatment of Bidirectional Typing

### 2024-01-18 tex/ESOP24/S3-defs.tex
Sprinkled the paper with commas; miscellaneous revisions

\begin{example} \label{ex:type-signature-for-function-type}
  Function types $A \bto B$ and typically a base type~$\mathtt{b}$ are included in simply typed $\lambda$-calculus and can be specified by the type signature $\Sigma_{\bto}$ consisting of operations $\mathsf{fun}$ and $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$. 
\end{example}

@@

A simply typed language specifies
\begin{inlineenum}
  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
\end{inlineenum}
Therefore, to specify a term construct, we enrich the notion of arity with some set of types for typing and extension context for variable binding.

\begin{defn}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
  A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$ where $T^*$ is the set of lists over $T$.
  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the \emph{target type}.
  For brevity, it is denoted by $\bargs \to A$ where $[\Delta_i]$ is omitted if empty.
\end{defn}

\begin{example}
  Observe that the $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
  \vspace{2em}
  \begin{equation*}
    \frac{\Gamma, \isTerm{x} : \eqnmarkbox[blue]{ext1}{A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
    \qquad\text{and}\qquad
    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
  \end{equation*}
  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension context}
  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument type}
  \annotate[yshift=0em]{below}{ty1,ty2}{target type}
  \vspace{.5em}
  \\
  if the empty context $\cdot$ is added verbosely, so they can be specified by arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$, respectively, with $\Type_{\Sigma_{\bto}}(A, B)$ as types.
\end{example}

@@

The definition of typing derivations is a bit more involved.
We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:

@@

\subsection{Bidirectional binding signatures and bidirectional type systems} \label{subsec:bidirectional-system}
For a bidirectional type system, typing judgements appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$, but these two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$, additionally indexed by a \emph{mode} $d : \Mode$---which can be either $\syn$ or $\chk$.
Therefore, to define a bidirectional type system, we enrich the concept of binding arity to \emph{bidirectional binding arity}, which further specifies the mode for each of its arguments and for the conclusion:

@@

\begin{defn}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Type_{\Sigma}(\Xi)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
  \]
\end{defn}
We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with $\Type_{\Sigma}(\Xi)$ as types.
We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.

\begin{defn}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  For a bidirectional type system $(\Sigma, \Omega)$,
  \begin{itemize}
    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
          \[
            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
          \]
          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
  \end{itemize}
  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
\end{defn}

@@

\begin{example}\label{ex:signature-simply-typed-lambda}
Having established generic definitions, we can now specify the simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of 
\[
  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
  \quad\text{and}\quad
  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
\]
\end{example}
More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.

>>>>

\begin{example} \label{ex:type-signature-for-function-type}
  Function types $A \bto B$ and typically a base type~$\mathtt{b}$ are included in simply typed $\lambda$-calculus, and can be specified by the type signature $\Sigma_{\bto}$ consisting of operations $\mathsf{fun}$ and $\mathsf{b}$ where $\arity(\mathsf{fun}) = 2$ and $\arity(\mathsf{b}) = 0$.
  Then, all types in simply typed $\lambda$-calculus can be given as $\Sigma_{\bto}$-types over the empty set, with $A \bto B$ introduced as $\tyOp_{\mathsf{fun}}(A, B)$ and $\mathtt{b}$ as $\tyOp_{\mathsf{b}}$. 
\end{example}

@@

A simply typed language specifies
\begin{inlineenum}
  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope), where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
\end{inlineenum}
Therefore, to specify a term construct, we enrich the notion of arity with some set of types for typing and extension context for variable binding.

\begin{defn}[{\cite[p.~322]{Fiore2010}}]\label{def:binding-arity}
  A \emph{binding arity} with a set $T$ of types is an inhabitant of $\left(T^* \times T\right)^* \times T$, where $T^*$ is the set of lists over $T$.
  In a binding arity $(((\Delta_1, A_1), \ldots, (\Delta_n, A_n)), A)$, every $\Delta_i$ and $A_i$ refers to the \emph{extension context} and the \emph{type of the $i$-th argument}, respectively, and $A$ the \emph{target type}.
  For brevity, it is denoted by $\bargs \to A$, where $[\Delta_i]$ is omitted if empty.
\end{defn}

\begin{example}
  Observe that the $\Rule{Abs}$ and $\Rule{App}$ rules in \Cref{fig:STLC-typing-derivations} can be read as
  \renewcommand{\eqnhighlightheight}{\vphantom{\Gamma}}
  \renewcommand{\eqnannotationfont}{\rmfamily\footnotesize} % package default
  \vspace{2em}
  \begin{equation*}
    \frac{\Gamma, \isTerm{x} : \eqnmarkbox[blue]{ext1}{A, \cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty1}{B}}{\Gamma \vdash \isTerm{\lam{x}t} : \eqnmarkbox{ty1}{A \bto B}}
    \qquad\text{and}\qquad
    \frac{\Gamma, \eqnmarkbox[blue]{ext2}{\cdot} \vdash \isTerm{t} : \eqnmarkbox[red]{aty2}{A \bto B} \qquad \Gamma, \eqnmarkbox[blue]{ext3}{\cdot} \vdash \isTerm{u} : \eqnmarkbox[red]{aty3}{A}}{\Gamma \vdash \isTerm{t\;u} : \eqnmarkbox{ty2}{B}}
  \end{equation*}
  \annotate[yshift=1.5em]{above}{ext1,ext2,ext3}{extension contexts}
  \annotate[yshift=0.4em]{above,right}{aty1,aty2,aty3}{argument types}
  \annotate[yshift=0em]{below}{ty1,ty2}{target types}
  \vspace{.5em}
  \\
  if the empty context $\cdot$ is added verbosely, so they can be specified by arities ${[A]B} \to {(A \bto B)}$ and ${(A \bto B), A} \to {B}$, respectively, with $\Type_{\Sigma_{\bto}}(A, B)$ as types.
\end{example}

@@

The definition of typing derivations is a bit more involved.
We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation of $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:

@@

\subsection{Bidirectional binding signatures and bidirectional type systems} \label{subsec:bidirectional-system}
Typing judgements for a bidirectional type system appear in two forms: $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
These two typing judgements can be considered as a single typing judgement $\Gamma |- t :^\dir{d} A$ indexed by a \emph{mode} $d : \Mode$, which can be either $\syn$ or $\chk$.
Therefore, to define a bidirectional type system, we enrich the concept of binding arity to \emph{bidirectional binding arity}, which further specifies the mode for each of its arguments and for the conclusion:

@@

\begin{defn}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Type_{\Sigma}(\Xi)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}\right)^* \times \Type_{\Sigma}(\Xi) \times {\Mode}.
  \]
\end{defn}
We write $\biop$ for an operation $o$ with a variable set $\Xi$ and its bidirectional binding arity with $\Type_{\Sigma}(\Xi)$ as types.
We call it \emph{checking} if $d$ is ${\chk}$ or \emph{synthesising} if $d$ is ${\syn}$; similarly, its $i$-th argument is checking if $d_i$ is $\chk$ and synthesising if $d_i$ is $\syn$.
A bidirectional type system $(\Sigma, \Omega)$ refers to a pair of a type signature $\Sigma$ and a bidirectional binding signature $\Omega$.

\begin{defn}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  For a bidirectional type system $(\Sigma, \Omega)$,
  \begin{itemize}
    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations}, and particularly
          \[
            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
          \]
          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
  \end{itemize}
  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
\end{defn}

@@

\begin{example}\label{ex:signature-simply-typed-lambda}
Having established generic definitions, we can now specify simply typed $\lambda$-calculus and its bidirectional type system---including raw terms, (bidirectional) typing derivations, and mode derivations---using just a pair of signatures $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and $\Omega^{\Leftrightarrow}_\Lambda$ which consists of 
\[
  \aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}
  \quad\text{and}\quad
  \aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}.
\]
\end{example}
More importantly, we are able to reason about constructions and properties that hold for any simply typed language with a bidirectional type system once and for all by quantifying over $(\Sigma, \Omega)$.

### 2024-01-18 tex/ESOP24/S4-pre-synthesis.tex
Sprinkled the paper with commas; miscellaneous revisions

Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
We discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
We formulate annotatability~\cite{Dunfield2021} and discuss its relationship with our completeness and generalised mode decoration in \cref{sec:annotatability}.

@@

Erasure of a bidirectional binding signature removes modes and keeps everything else intact; this can be straightforwardly extended by induction to remove modes from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
Instead we can remove typing and retain modes, arriving at a mode derivation.
Conversely, if we have both mode and typing derivations for the same term, we can combine them to form a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are merely the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
All these can be summarised in one theorem and proved by induction.

@@

This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules: having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts: every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
With strictness coming into play, non-genuine derivations are now `truly non-genuine'.

@@

\subsection{Annotatability}\label{sec:annotatability}
\citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
In our theory, we may formulate annotatability as follows.

>>>>

Our first important construction is mode decoration in \cref{sec:mode-decoration}, which is in fact generalised to pinpoint any missing type annotations in a given raw term.
We discuss some related properties: by bringing mode derivations into the picture, we are able to give a natural formulation of soundness and completeness of a bidirectional type system with respect to its erasure to an ordinary type system in \cref{sec:soundness-and-completeness}.
We also reformulate annotatability~\cite[Section~3.2]{Dunfield2021} and compare it with our completeness and generalised mode decoration in \cref{sec:annotatability}.

@@

Erasure of a bidirectional binding signature removes modes and keeps everything else intact; this can be straightforwardly extended by induction to remove modes from a bidirectional typing derivation and arrive at an ordinary typing derivation, which is soundness.
Alternatively, we can remove typing and retain modes, arriving at a mode derivation.
Conversely, if we have both mode and typing derivations for the same term, we can combine them to form a bidirectional typing derivation, which is completeness.
In short, soundness and completeness are merely the separation and combination of mode and typing information carried by the three kinds of derivations while keeping their basic structure, directed by the same raw term.
All these can be summarised in one theorem and proved by induction.

@@

This lemma would be wrong if the `strictness' boolean~$s$ was left out of the rules: having both $\ChkRule{Sub}$ and $\SynRule{Missing}$, which we call \emph{mode casts}, it would be possible to switch between the two modes freely, which unfortunately means that we could insert a pair of $\ChkRule{Sub}$ and $\SynRule{Missing}$ anywhere, constructing a non-genuine derivation even when there is in fact a genuine one.
The `strictness' boolean~$s$ can be thought of as disrupting the formation of such pairs of mode casts: every rule other than the mode casts sets~$s$ to~$\BTrue$, meaning that a term is \emph{strictly} in the mode assigned by the rule (i.e.~not altered by a mode cast), whereas the mode casts set~$s$ to~$\BFalse$.
Furthermore, the sub-derivation of a mode cast has to be strict, so it is impossible to have consecutive mode casts.
Another way to understand the role of~$s$ is that it makes the $\SynRule{Missing}$ rule precise: an annotation is truly missing only when a term is \emph{strictly} in checking mode but is required to be in synthesising mode.
The explicit formulation of strictness makes non-genuine derivations `truly non-genuine', and \cref{thm:Pre?-false} can be proved.

@@

\subsection{Annotatability}\label{sec:annotatability}
\citet[Section~3.2]{Dunfield2021} formulated completeness differently from our \cref{lem:soundness-completeness} and proposed \emph{annotatability} as a more suitable name.
In our theory, we may reformulate annotatability as follows.
